Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-22 paper/paper.tex
edits on Appendix

\section{The reason to not using the standard tensor network notations}\label{app:tensorbad}
As we have mentioned in the main text,
a standard tensor network notation is equivalent to the generalized tensor network by introducing $\delta$ tensors,
where a $\delta$ tensor of rank $d$ is defined as
\begin{equation}
    \delta_{i_1, i_2,\ldots,i_d} = \begin{cases}
        1, & i_1=i_2=\ldots =i_d,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}

@@

In this diagram, the circle on each vertex in the original graph is a $\delta$ tensor of rank $8$.
If we contract this tensor network in a naive column-wise order, the maximum intermediate tensor has rank $\sim3L$, requiring a storage of size $\approx 2^{3L}$.
If we relax the restriction that each label appears exactly twice. We have the following hypergraph representation of a generalized tensor network.

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \def\sql{0.24}
    \pgfmathsetseed{2}
    \foreach[evaluate={\cr=0.1+0.5*Mod(\x,2)}] \x in {1,...,5}
        \foreach[evaluate={\cg=0.1+0.3*Mod(\y,2); \cy=0.5-0.5*Mod(\y,2)}] \y in {1,...,5}{
            \edef\R{\pdfuniformdeviate 255}
            \edef\G{\pdfuniformdeviate 255}
            \edef\B{\pdfuniformdeviate 255}
            \xdefinecolor{MyColor}{RGB}{\R,\G,\B}
            \ifnum \x < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y+0.3);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y-0.3);
                \fi
            \fi
            \ifnum \x > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y+0.7);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y-0.7);
                \fi
            \fi
            \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y+0.5);
            \fi
            \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y-0.5);
            \fi
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            %\filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
                \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \fi
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
Here, we use different colors to distinguish different hyperedges.
A vertex tensor always has rank $1$ and is not shown here since it does not change the contraction complexity.
Again, if we contract this tensor network in the column-wise order, the maximum intermediate tensor rank is $\sim L$, which can be see by counting the number of colors.

\section{Generalizing to other graph problems}\label{app:otherproblems}

@@

which can help us understand why the program for solving MIS is trapped in a local minimum.

@@

Comparing with the independence polynomial in \Eq{eq:idpdef}, we have $b_{k} \leq a_{k}$ and $b_{\alpha(G)} = a_{\alpha(G)}$. $I_{\rm max}(G, 1)$ counts the total number of maximal independent sets~\cite{Gaspers2012, Manne2013}, where the fastest algorithm currently has a runtime of $O(1.3642^{|V|})$~\cite{Gaspers2012}.

@@

By contracting this tensor network with generic element type,
we can compute the maximum independent set properties such as maximal independence polynomial, enumerating maximal independent sets.
Let us consider the example in~\Sec{eg:tensorcontraction}: its corresponding tensor network structure for computing the maximal independent polynomial becomes

@@

One can see that the average degree of a tensor is increased.
The computational complexity of this new tensor network contraction is often greater than the one for computing the independence polynomial.
However, for most sparse graphs, this tensor network contraction approach is still much faster than enumerating all the maximal cliques on its complement graph using the Bron-Kerbosch algorithm~\cite{Bron1973}, which is the standard algorithm that we are aware of to compute the maximal independence polynomial.
We show the benchmark of computing the maximal independent set properties in \Fig{fig:benchmark-maximal},
including a comparison to the Bron-Kerbosch algorithm from Julia package Graphs~\cite{Graphs}.
%Since the network for computing maximal independent set properties is different from the one for computing the independence polynomial,
the treewidth of this tensor network is significantly larger, hence only small graphs can be benchmarked.
The time for the tensor network approach and the Bron-Kerbosch approach to enumerate all maximal independent sets are comparable,
while the tensor network does counting much more efficiently.
Due to the memory limit, this Bron-Kerbosch algorithm stops working at size $70$ and above.

@@

\subsection{Matching problem}
A matching polynomial of a graph $G$ is defined as
\begin{equation}
    M(G, x) = \sum\limits_{k=1}^{|V|/2} c_k x^k,
\end{equation}
where $k$ is the number of matches, and coefficients $c_k$ are the corresponding counting.
We map an edge $(u, v) \in E$ to a label $\langle u, v\rangle \in \{0, 1\}$ in a tensor network,
where $1$ means two vertices of an edge are matched, $0$ means otherwise.
Then we define a tensor of rank $d(v) = |N(v)|$ on vertex $v$ such that,
\begin{equation}
    W_{\langle v, n_1\rangle, \langle v, n_2 \rangle, \ldots, \langle v, n_{d(v)}\rangle} = \begin{cases}
        1, & \sum_{i=1}^{d(v)} \langle v, n_i \rangle \leq 1,\\
        0, & \text{otherwise},
    \end{cases}
\end{equation}
and a tensor of rank $1$ on the bond
\begin{equation}
    B_{\langle v, w\rangle} = \begin{cases}
    1, & \langle v, w \rangle = 0 \\
    x, & \langle v, w \rangle = 1,
\end{cases}
\end{equation}
where label $\langle v, w \rangle$ is equivalent to $\langle w,v\rangle$.
Here, a vertex tensor specifies the restriction that a vertex can not be in two matched edges, while an edge tensor contributes the variable in the polynomial.

\subsection{k-Colouring}
Let us use 3-colouring problem defined on vertices as an example.
For a vertex $v$, we define the degree of freedoms $c_v\in\{1,2,3\}$ and a vertex tensor labelled by it as
\begin{equation}
    W(v) = \left(\begin{matrix}
        r_v\\
        g_v\\
        b_v
    \end{matrix}\right).
\end{equation}
For an edge $(u, v)$, we define an edge tensor as a matrix labelled by $(c_u, c_v)$ to specify the constraint
\begin{equation}
    B = \left(\begin{matrix}
        0 & 1 & 1\\
        1 & 0 & 1\\
        1 & 1 & 0
    \end{matrix}\right).
\end{equation}
The number of possible colouring can be obtained by contracting this tensor network by setting vertex tensor elements $r_v, g_v$ and $b_v$ to $1$.
By designing generic types as tensor elements, one can get other properties.
Similarly, one can define the k-colouring problem on edges too by switching the roles of edges and vertices.

\subsection{Max cut problem}
Max cut problem is also known as the boolean spin glass problem.
For a vertex $v\in V$, we define a boolean degree of freedom $s_v\in\{0, 1\}$.
Then the max cut problem can be encoded to tensor networks by mapping an edge $(i,j)\in E$ to an edge matrix labelled by $s_is_j$
\begin{equation}
    B(x_{\langle i, j\rangle}) = \left(\begin{matrix}
        1 & x_{\langle i, j\rangle}\\
        x_{\langle i, j\rangle} & 1
    \end{matrix}\right),
\end{equation}
where variable $x_{\langle i, j\rangle}$ represents a cut on edge $(i, j)$ or a domain wall of an Ising spin glass.
Similar to other problems, we can define a polynomial about edges variables by setting $x_{\langle i, j\rangle} = x$,
where its $k$th coefficient is two times the number of configurations of cut size $k$.

\subsection{Set packing}
Set packing is the hypergraph generalization of the maximum independent set problem, where a set corresponds to a vertex and an element corresponds to a hyperedge.
To solve the set packing problem, we just remove the rank 2 restriction of the edge tensor in \Eq{eq:edgetensor}
\begin{equation}
    B_{v,w,\ldots, z} = \begin{cases}
        1, & v+w+\ldots+z\leq 1,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}

@@

\section{Integer sequences formed by the number of independent sets}

We computed the number of independent sets on square lattices and King's graphs
with our generic tensor network contraction on GPUs.
The tensor element type is finite field algebra so that we can reach arbitrary precision.
We also computed independence polynomial rigorously for these lattices in our \href{https://github.com/GiggleLiu/NoteOnTropicalMIS/tree/master/data}{Github repo}.

\begin{table}[h]
\caption{The number of independent sets for square grid graphs of size $L\times L$. This forms the integer sequence \href{https://oeis.org/A006506}{OEIS A006506}.
Here we only show two updated entries for $L=38,39$, which to our knowledge, has not been computed before.~\cite{Butera2014}
}
\begin{center}
\scalebox{0.9}{
\begin{tabular}{|c| >{\centering\arraybackslash} p{0.95\linewidth}|}
 \hline $L$  & square grid graphs \\
 \hline $38$ & 616 412 251 028 728 207 385 738 562 656 236 093 713 609 747 387 533 907 560 081 990 229 746 115 948 572 583 817 557 035 128 726 922 565 913 748 716 778 414 190 432 479 964 245 067 083 441 583 742 870 993 696 157 129 887 194 203 643 048 435 362 875 885 498 554 979 326 352 127 528 330 481 118 313 702 375 541 902 300 956 879 563 063 343 972 979\\
 \hline $39$ &  29 855 612 447 544 274 159 031 389 813 027 239 335 497 014 990 491 494 036 487 199 167 155 042 005 286 230 480 609 472 592 158 583 920 411 213 748 368 073 011 775 053 878 033 685 239 323 444 700 725 664 632 236 525 923 258 394 737 964 155 747 730 125 966 370 906 864 022 395 459 136 352 378 231 301 643 917 282 836 792 261 715 266 731 741 625 623 207 330 411 607\\
  \hline
\end{tabular}
}
\end{center}
\label{tbl:squaregrid}
\end{table}
% online digit seperator: https://www.browserling.com/tools/thousands-separator

>>>>

\section{Additional contraction overhead using the standard tensor network notations}\label{app:tensorbad}
As we have mentioned in the main text,
a standard tensor network notation is equivalent to the generalized tensor network by introducing $\delta$ tensors,
where a $\delta$ tensor of rank $d$ is defined as
\begin{equation}
    \delta_{i_1, i_2,\ldots,i_d} = \begin{cases}
        1, & i_1=i_2=\ldots =i_d,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}

@@

In this diagram, the circle on each vertex in the original graph is a $\delta$ tensor up to rank $8$.
If we contract this tensor network in a naive column-wise order, the maximum intermediate tensor has rank $\sim3L$, requiring a storage of size $\approx 2^{3L}$.
If we relax the restriction that each label must appears exactly twice, we have the following hypergraph representation of a generalized tensor network.

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \def\sql{0.24}
    \pgfmathsetseed{2}
    \foreach[evaluate={\cr=0.1+0.5*Mod(\x,2)}] \x in {1,...,5}
        \foreach[evaluate={\cg=0.1+0.3*Mod(\y,2); \cy=0.5-0.5*Mod(\y,2)}] \y in {1,...,5}{
            \edef\R{\pdfuniformdeviate 255}
            \edef\G{\pdfuniformdeviate 255}
            \edef\B{\pdfuniformdeviate 255}
            \xdefinecolor{MyColor}{RGB}{\R,\G,\B}
            \ifnum \x < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y+0.3);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y-0.3);
                \fi
            \fi
            \ifnum \x > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y+0.7);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y-0.7);
                \fi
            \fi
            \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y+0.5);
            \fi
            \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y-0.5);
            \fi
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            %\filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
                \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \fi
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
Here, we use different colors to distinguish different hyperedges.
A vertex tensor always has rank $1$ and is not shown here since it does not change the contraction complexity.
If we contract this tensor network in the column-wise order, the maximum intermediate tensor rank is $\sim L$, which can be seen by counting the number of colors.

\section{Generalization to other graph problems}\label{app:otherproblems}

@@

which can be helpful for understanding why the program for solving MIS can be trapped in a local minimum.

@@

Comparing with the independence polynomial in \Eq{eq:idpdef}, we have $b_{k} \leq a_{k}$ and $b_{\alpha(G)} = a_{\alpha(G)}$. $I_{\rm max}(G, 1)$ counts the total number of maximal independent sets~\cite{Gaspers2012, Manne2013}, where, to our knowledge, the fastest algorithm currently has a runtime of $O(1.3642^{|V|})$~\cite{Gaspers2012}.

@@

By contracting this tensor network with generic element types,
we can compute the maximal independent set properties such as the maximal independence polynomial and the enumeration of maximal independent sets.
Let us consider the example in~\Sec{eg:tensorcontraction}: its corresponding tensor network structure for computing the maximal independent polynomial becomes

@@

One can see that the average degree of a tensor is increased.
The computational complexity of this new tensor network contraction is often greater than the one for computing the independence polynomial.
However, for most sparse graphs, this tensor network contraction approach is still much faster than enumerating all the maximal cliques on its complement graph using the Bron-Kerbosch algorithm~\cite{Bron1973}, which is the standard algorithm that we are aware of to compute the maximal independence polynomial.
We show the benchmark of computing the maximal independent set properties in \Fig{fig:benchmark-maximal},
including a comparison to the Bron-Kerbosch algorithm from Julia package Graphs~\cite{Graphs}.
%Since the network for computing maximal independent set properties is different from the one for computing the independence polynomial,
The treewidth of this tensor network is significantly larger, so benchmark with only a smaller graph size is feasible.
The time for the tensor network approach and the Bron-Kerbosch approach to enumerate all maximal independent sets are comparable,
while the tensor network does counting much more efficiently.
Due to the memory limit, this Bron-Kerbosch algorithm is only feasible up to a graph size around $70$.

@@

\subsection{Matching problem}
A matching polynomial of a graph $G$ is defined as
\begin{equation}
    M(G, x) = \sum\limits_{k=1}^{|V|/2} c_k x^k,
\end{equation}
where $k$ is the size of a matching, and coefficients $c_k$ are the corresponding counting of the number of matchings of the size $k$.
We map an edge $(u, v) \in E$ to a label $\langle u, v\rangle \in \{0, 1\}$ in a tensor network,
where $1$ means two vertices of an edge are matched and $0$ otherwise.
Then we define a tensor of rank $d(v) = |N(v)|$ on vertex $v$ such that,
\begin{equation}
    W_{\langle v, n_1\rangle, \langle v, n_2 \rangle, \ldots, \langle v, n_{d(v)}\rangle} = \begin{cases}
        1, & \sum_{i=1}^{d(v)} \langle v, n_i \rangle \leq 1,\\
        0, & \text{otherwise},
    \end{cases}
\end{equation}
and a tensor of rank $1$ on the bond
\begin{equation}
    B_{\langle v, w\rangle} = \begin{cases}
    1, & \langle v, w \rangle = 0 \\
    x, & \langle v, w \rangle = 1,
\end{cases}
\end{equation}
where label $\langle v, w \rangle$ is equivalent to $\langle w,v\rangle$.
Here, a vertex tensor specifies the restriction that a vertex cannot be matched by more than one edge, while an edge tensor contributes the variable in the polynomial.

\subsection{$k$-Coloring}
Let us use the 3-coloring problem defined on vertices as an example.
For a vertex $v$, we define the degrees of freedom $c_v\in\{1,2,3\}$ and a vertex tensor labelled by it as
\begin{equation}
    W(v) = \left(\begin{matrix}
        r_v\\
        g_v\\
        b_v
    \end{matrix}\right).
\end{equation}
For an edge $(u, v)$, we define an edge tensor as a matrix labelled by $(c_u, c_v)$ to specify the constraint
\begin{equation}
    B = \left(\begin{matrix}
        0 & 1 & 1\\
        1 & 0 & 1\\
        1 & 1 & 0
    \end{matrix}\right).
\end{equation}
The number of possible colorings can be obtained by contracting this tensor network by setting vertex tensor elements $r_v, g_v$ and $b_v$ to $1$.
By designing generic types as tensor elements, one can get other properties.
Similarly, one can also define the $k$-coloring problem on edges by switching the roles of edges and vertices.

\subsection{Max-Cut problem}
The Max-Cut problem is also known as the boolean spin glass problem.
For a vertex $v\in V$, we define a boolean degree of freedom $s_v\in\{0, 1\}$.
Then the Max-Cut problem can be encoded to tensor networks by mapping an edge $(i,j)\in E$ to an edge matrix labelled by $s_is_j$
\begin{equation}
    B(x_{\langle i, j\rangle}) = \left(\begin{matrix}
        1 & x_{\langle i, j\rangle}\\
        x_{\langle i, j\rangle} & 1
    \end{matrix}\right),
\end{equation}
where variable $x_{\langle i, j\rangle}$ represents a cut on edge $(i, j)$ or a domain wall of an Ising spin glass.
Similar to other problems, we can define a polynomial with edge variables by setting $x_{\langle i, j\rangle} = x$,
where its $k$th coefficient is two times the number of configurations of cut size $k$.

\subsection{Set packing}
Set packing is the hypergraph generalization of the maximum independent set problem, where a set corresponds to a vertex and an element corresponds to a hyperedge.
To solve the set packing problem, we can just remove the rank-2 restriction of the edge tensor in \Eq{eq:edgetensor}
\begin{equation}
    B_{v,w,\ldots, z} = \begin{cases}
        1, & v+w+\ldots+z\leq 1,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}

@@

\section{Integer sequence formed by the number of independent sets}

We computed the number of independent sets on square lattices and King's graphs
with our generic tensor network contraction algorithm on GPUs.
The tensor element type is finite-field algebra so that we can reach arbitrary precision.
We also computed the independence polynomial exactly for these lattices which can be found in our \href{https://github.com/GiggleLiu/NoteOnTropicalMIS/tree/master/data}{Github repo}.

\begin{table}[h]
\caption{The number of independent sets for square grid graphs of size $L\times L$. This forms the integer sequence \href{https://oeis.org/A006506}{OEIS A006506}.
Here we only include two updated entries for $L=38,39$, which, to our knowledge, has not been computed before~\cite{Butera2014}.
}
\begin{center}
\scalebox{0.9}{
\begin{tabular}{|c| >{\centering\arraybackslash} p{0.95\linewidth}|}
 \hline $L$  & square grid graphs \\
 \hline $38$ & 616 412 251 028 728 207 385 738 562 656 236 093 713 609 747 387 533 907 560 081 990 229 746 115 948 572 583 817 557 035 128 726 922 565 913 748 716 778 414 190 432 479 964 245 067 083 441 583 742 870 993 696 157 129 887 194 203 643 048 435 362 875 885 498 554 979 326 352 127 528 330 481 118 313 702 375 541 902 300 956 879 563 063 343 972 979\\
 \hline $39$ &  29 855 612 447 544 274 159 031 389 813 027 239 335 497 014 990 491 494 036 487 199 167 155 042 005 286 230 480 609 472 592 158 583 920 411 213 748 368 073 011 775 053 878 033 685 239 323 444 700 725 664 632 236 525 923 258 394 737 964 155 747 730 125 966 370 906 864 022 395 459 136 352 378 231 301 643 917 282 836 792 261 715 266 731 741 625 623 207 330 411 607\\
  \hline
\end{tabular}
}
\end{center}
\label{tbl:squaregrid}
\end{table}
% online digit seperator: https://www.browserling.com/tools/thousands-separator
